Testable normative theories

### 2020-04-15 simple.tex
Switched to new canvas

\RequirePackage[l2tabu, orthodox]{nag}
\documentclass[version=last, pagesize, twoside=off, bibliography=totoc, DIV=calc, fontsize=14pt, a4paper, french, english]{scrartcl}
\input{preamble/packages}
\input{preamble/math_basics}
\input{preamble/math_mine}
\input{preamble/redac}
\input{preamble/draw}
\input{preamble/acronyms}
\addtokomafont{labelinglabel}{\sffamily\bfseries}
\renewcommand{\phi}{\varphi}

@@

\begin{example}
	Consider a decision problem where a set of alternatives $\allalts$ is given, containing food products among which $i$ would like to form a deliberated preference (taking into account the effects of food on health, price, pleasure, morality issues related to the production process, and so on). The topic could be defined as $\alltopic$ = $\Set{\prop_\alt, ¬\prop_\alt, \forall \alt \in \allalts}$, where $\prop_\alt$ is the proposition according to which $\alt$ is one of the best alternatives among $\allalts$ (there is no strictly better alternatives among $\allalts$, from $i$’s point of view), and $¬\prop_\alt$ is the proposition according to which $\alt$ is not one of the best alternatives among $\allalts$ (some $\alt' \in \allalts$ is a strictly better alternative, from $i$’s point of view). 
	The set $\allargs$ is defined as the set of all strings, because in this example it is considered acceptable to restrict arguments to those that can take a textual form.
\end{example}

@@

It is also necessary to capture the evolution of the position of $i$ towards arguments over time: a repeated query using a given pair of arguments may yield different answers, for example because the repeated question has been interleaved with another query containing other arguments (which $i$ may still have in mind when answering the second repetition), or for any other known or unknown reason. The important information for us is the \emph{set} of answers that $i$ could give to a query involving two given arguments and a given topic, among the set $\{\prop, ¬\prop, \text{both}\}$. The set designates all possible answers over time and over different order of asking the queries.
Hence, given $\prop \in \alltopic$, define a relation $\ileadsto^\prop \subseteq (\allargs × \allargs) × \powersetz{\{\prop, ¬\prop, \text{both}\}}$, where $\powersetz{B}$ designates the subsets of $B$ excluding the emptyset, with the semantics that $(\ar, \ar') \ileadsto^\prop C$, with $C \subseteq \Set{\prop, ¬\prop, \text{both}}$, iff $\forall c \in C$, it is observable at least once that $i$, when presented with $(\ar, \ar')$, considers $c$ valid in her current state of mind. The querying protocol is supposedly made so that for any $\prop \in \alltopic$, $\ileadsto^\prop = \ileadsto^{¬\prop}$.

This information is only indirectly observable: for example, if $(\ar, \ar') \ileadsto^\prop \{\prop\}$, a querier will never know more than $(\ar, \ar') \ileadsto^\prop C$ with $\prop \in C \subseteq \Set{\prop, ¬\prop, \text{both}}$. 
Also, it may be only possible to test one ordering of the queries (assuming that $i$ forgets over time might permit to test different orderings, but this assumption might be unrealistic). In some cases, it may even be only possible to test a single query with a given individual.
This limitation of our observation will not be a problem for the results of this article. Intuitively, this bears on the fact that, first, theories apply to multiple individuals, and second, if a theory claims that $(\ar, \ar') \ileadsto^\prop \{\prop\}$, then either this claim is correct, or it is falsifiable, in the sense that there is a possibility of observing a contradiction to its claim.

@@

Define $\ileadstost$ from $\ileadsto^\prop$ as follows, for any $\prop \in \alltopic$: $(\ar, \ar') \ileadstost (\prop, \text{possible}) ⇔ (\ar, \ar') \ileadsto^\prop C$ with $C \subseteq \Set{\prop, \text{both}}$; and $(\ar, \ar') \ileadstost (\prop, \text{sure}) ⇔ (\ar, \ar') \ileadsto^\prop \set{\prop}$.

@@

\subsection{Deliberated preference}
The following important definition is required in order to define the deliberated preference of $i$. An argument $\ar$ is decisive for $\phi$, in short, $\phi\text{-dec}$ iff $\forall \ar' \in \allargs: (\ar', \ar) \ileadstost \phi$. 
\footnote{Equivalently: $\ar \ \phidec ⇔ [\ileadstostinv(\phi)](\ar) = \allargs$, and $\exists \ar \ \phidec ⇔ [\ileadstostinv(\phi)]^{-1}(\allargs) ≠ \emptyset$.}
I also say that $\ar$ is $\propsuredec$, $\proppossdec$, or $¬\proppossdec$, when $\phi=(\prop, \text{sure})$, $\phi=(\prop, \text{possible})$, or $\phi=(¬\prop, \text{possible})$, respectively.
\footnote{$\propsuredec ⇔ \forall \ar': (\ar', \ar) \ileadstosst \prop ⇔ \forall \ar': ¬[(\ar', \ar) \ileadstope ¬\prop]$, $\proppossdec$ ⇔ $\forall \ar': (\ar', \ar) \ileadstopst \prop$. Note that “both” ≠ “we donno”: in case “both”, we know how to prove that both are reasonable options, in the sense that neither is certain. We have excluded some possibilities (and some other model could disagree with the “both” conclusion).}

@@

@@

\section{Model certainties}
Assume we define $\ar_1 \wibeatse[¬\prop, \text{sure}] \ar_2$ ⇔ $\ar_2 \nwibeatse[\prop] \ar_1$. Then, indeed, given $\ar_1 \mleadsto \prop$, $\ar_2 \wibeatse[¬\prop, \text{sure}] \ar_1$ ⇔ $\ar_2 \ibeatse[¬\prop, \text{sure}] \ar_1$. But it gives the wrong conclusion. For $\ar_3 \mbeats \ar_2 \mbeats \ar_1 \mleadsto t: \ar_3 \wibeatse[\prop, \text{sure}] \ar_2$ iff $\ar_2 \nwibeatse[¬\prop] \ar_3$ iff $\exists \ar_1 \in \mbeats(\ar_3) \suchthat \ar_3 \wibeatse[¬\prop] \ar_1 ∧ \ar_3 \wibeatse[¬\prop] \ar_2 + \ar_1$.

>>>>

\RequirePackage[l2tabu, orthodox]{nag}
\documentclass[version=last, pagesize, twoside=off, bibliography=totoc, DIV=calc, fontsize=12pt, a4paper, french, english]{scrartcl}
\input{preamble/packages}
\input{preamble/math_basics}
\input{preamble/math_mine}
\input{preamble/redac}
%\input{preamble/draw}
%\input{preamble/acronyms}
\addtokomafont{labelinglabel}{\sffamily\bfseries}
\renewcommand{\phi}{\varphi}

%I find these settings useful in draft mode. Should be removed for final versions.
	%Which line breaks are chosen: accept worse lines, therefore reducing risk of overfull lines. Default = 200.
		\tolerance=2000
	%Accept overfull hbox up to...
		\hfuzz=2cm
	%Reduces verbosity about the bad line breaks.
		\hbadness 5000
	%Reduces verbosity about the underful vboxes.
		\vbadness=1300

@@

\begin{example}
	Consider a decision problem where a set of alternatives $\allalts$ is given, containing food products among which $i$ would like to form a deliberated preference (taking into account the effects of food on health, price, pleasure, morality issues related to the production process, and so on). The topic could be defined as $\alltopic$ = $\set{\prop_\alt, ¬\prop_\alt, \forall \alt \in \allalts}$, where $\prop_\alt$ is the proposition according to which $\alt$ is one of the best alternatives among $\allalts$ (there is no strictly better alternatives among $\allalts$, from $i$’s point of view), and $¬\prop_\alt$ is the proposition according to which $\alt$ is not one of the best alternatives among $\allalts$ (some $\alt' \in \allalts$ is a strictly better alternative, from $i$’s point of view). 
	The set $\allargs$ is defined as the set of all strings, because in this example it is considered acceptable to restrict arguments to those that can take a textual form.
\end{example}

@@

It is also necessary to capture the evolution of the position of $i$ towards arguments over time: a repeated query using a given pair of arguments may yield different answers, for example because the repeated question has been interleaved with another query containing other arguments (which $i$ may still have in mind when answering the second repetition), or for any other known or unknown reason. The important information for us is the \emph{set} of answers that $i$ could give to a query involving two given arguments and a given topic, among the set $\{\prop, ¬\prop, \text{both}\}$. The set designates all possible answers over time and over different order of asking the queries.
Hence, given $\prop \in \alltopic$, define a relation $\ileadsto^\prop \subseteq (\allargs × \allargs) × \powersetz{\{\prop, ¬\prop, \text{both}\}}$, where $\powersetz{B}$ designates the subsets of $B$ excluding the emptyset, with the semantics that $(\ar, \ar') \ileadsto^\prop C$, with $C \subseteq \set{\prop, ¬\prop, \text{both}}$, iff $\forall c \in C$, it is observable at least once that $i$, when presented with $(\ar, \ar')$, considers $c$ valid in her current state of mind. The querying protocol is supposedly made so that for any $\prop \in \alltopic$, $\ileadsto^\prop = \ileadsto^{¬\prop}$.

This information is only indirectly observable: for example, if $(\ar, \ar') \ileadsto^\prop \{\prop\}$, a querier will never know more than $(\ar, \ar') \ileadsto^\prop C$ with $\prop \in C \subseteq \set{\prop, ¬\prop, \text{both}}$. 
Also, it may be only possible to test one ordering of the queries (assuming that $i$ forgets over time might permit to test different orderings, but this assumption might be unrealistic). In some cases, it may even be only possible to test a single query with a given individual.
This limitation of our observation will not be a problem for the results of this article. Intuitively, this bears on the fact that, first, theories apply to multiple individuals, and second, if a theory claims that $(\ar, \ar') \ileadsto^\prop \{\prop\}$, then either this claim is correct, or it is falsifiable, in the sense that there is a possibility of observing a contradiction to its claim.

@@

Define $\ileadstost$ from $\ileadsto^\prop$ as follows, for any $\prop \in \alltopic$: $(\ar, \ar') \ileadstost (\prop, \text{possible}) ⇔ (\ar, \ar') \ileadsto^\prop C$ with $C \subseteq \set{\prop, \text{both}}$; and $(\ar, \ar') \ileadstost (\prop, \text{sure}) ⇔ (\ar, \ar') \ileadsto^\prop \set{\prop}$.

@@

\subsection{Deliberated preference}
The following important definition is required in order to define the deliberated preference of $i$. An argument $\ar$ is decisive for $\phi$, in short, $\phi\text{-dec}$ iff $\forall \ar' \in \allargs: (\ar', \ar) \ileadstost \phi$. 
\footnote{Equivalently: $\ar \ \phidec ⇔ [\ileadstostinv(\phi)](\ar) = \allargs$, and $\exists \ar \ \phidec ⇔ [\ileadstostinv(\phi)]^{-1}(\allargs) ≠ \emptyset$.}
I also say that $\ar$ is $\propsuredec$, $\proppossdec$, or $¬\proppossdec$, when $\phi=(\prop, \text{sure})$, $\phi=(\prop, \text{possible})$, or $\phi=(¬\prop, \text{possible})$, respectively.
\footnote{$\propsuredec ⇔ \forall \ar': (\ar', \ar) \ileadstosst \prop ⇔ \forall \ar': ¬[(\ar', \ar) \ileadstope ¬\prop]$, $\proppossdec ⇔ \forall \ar': (\ar', \ar) \ileadstopst \prop$. Note that “both” $≠$ “we donno”: in case “both”, we know how to prove that both are reasonable options, in the sense that neither is certain. We have excluded some possibilities (and some other model could disagree with the “both” conclusion).}

@@

\bibliography{simple}
\end{document}

@@

\section{Model certainties}
Assume we define $\ar_1 \wibeatse[¬\prop, \text{sure}] \ar_2 ⇔ \ar_2 \nwibeatse[\prop] \ar_1$. Then, indeed, given $\ar_1 \mleadsto \prop$, $\ar_2 \wibeatse[¬\prop, \text{sure}] \ar_1 ⇔ \ar_2 \ibeatse[¬\prop, \text{sure}] \ar_1$. But it gives the wrong conclusion. For $\ar_3 \mbeats \ar_2 \mbeats \ar_1 \mleadsto t: \ar_3 \wibeatse[\prop, \text{sure}] \ar_2$ iff $\ar_2 \nwibeatse[¬\prop] \ar_3$ iff $\exists \ar_1 \in \mbeats(\ar_3) \suchthat \ar_3 \wibeatse[¬\prop] \ar_1 ∧ \ar_3 \wibeatse[¬\prop] \ar_2 + \ar_1$.
